A highly-available move operation for replicated trees

### 2019-06-19 paper/move-op.tex
Updates to related work.

Isabelle/HOL can generate executable Haskell, OCaml, Scala, and Standard ML code from HOL type and constant definitions using a sophisticated code generation mechanism~\cite{DBLP:conf/flops/HaftmannN10}.
However,  care must be taken when using this mechanism, as:
\begin{enumerate}
\item
The mechanism is \emph{partial}: not all HOL definitions can be realised in executable form.
Definitions making essential use of choice principles (e.g.~our \isa{get\_parent}) are obvious examples of definitions that will cause problems.
However, \emph{inductively defined} relations characterised by a series of introduction rules, such as our \isa{ancestor} relation, are also problematic as their execution necessarily requires Prolog-like proof search.
\item
Na\"{i}ve use often leads to inefficient target language code, especially true when using HOL's \isa{set} type, a type of potentially infinite sets.
Whilst convenient for theorem proving this type often leads to inefficiency in generated code.
\end{enumerate}

We therefore produce a series of definitions that are optimised for execution rather than theorem proving, before trying to evaluate our algorithms.
Notably, we eschew the HOL \isa{set} type, and instead use a more efficient hash-map implementation provided by the Isabelle Collections Framework~\cite{DBLP:conf/itp/LammichL10}, a library of verified, efficient container datatypes and associated algorithms.

Write $('a::\{hashable\}, 'b)\ \isa{hm}$ for the type of hash-maps with key type $\isa{'a}$ and binding type $\isa{'b}$.
Note that keys have an explicit $\isa{hashable}$ constraint, meaning that this hash-map type may only be used with kay types with an associated hashing function.
With this type, we represent trees by the type $(\isa{'n}::\{hashable\}, \isa{'m} \times \isa{'n})\ \isa{hm}$, mapping from a \emph{child} node to its \emph{parent} node and associated metadata, if any such mapping exists.

We use a \emph{simulation relation} between our two representations of trees, and say that a hash-map $\isa{t}$ of type $(\isa{'n}, \isa{'m} \times \isa{'n})\ \isa{hm}$ \emph{simulates} a set $T$ of type $(\isa{'n} \times \isa{'m} \times \isa{'n})\ \isa{set}$ whenever
\begin{displaymath}
\isa{hm.lookup}\ c\ t = \isa{Some}\ (m, p) \quad\text{iff}\quad (p, m, c) \in T \quad\text{forall $c$, $m$, and $p$}
\end{displaymath}
where $hm.lookup$ searches for a binding for key $\isa{c}$ in hash-map $t$, returning $\isa{Some}\ (m, p)$ when a binding from $c$ to $(m, p)$ exists in the hash-map, or $\isa{None}$ otherwise.
Write $t \preceq T$ to assert that $t$ and $T$ are contained within the simulation relation.
Note some corollaries follow immediately: e,g,~we have $t \preceq T$ implies $\isa{unique\_parent}\ \isa{T}$.

Our approach is to produce a new set of definitions---mirroring the structure of our existing ones---wherein we replace all occurrences of the $(\isa{'n} \times \isa{'m} \times \isa{'n})\ \isa{set}$ type, and any associated set operations, with the $(\isa{'n}, \isa{'m} \times \isa{'n})\ \isa{hm}$ type and associated hash-map operations.
For clarity, we refer to any definition using hash-maps as an ``efficient'' definition, and any definition making use of the \isa{set} type as an ``abstract'' definition.
For most definitions this is a mechanical process, and we elide further discussion of the required changes, referring the interested reader to our Isabelle/HOL source code.
However \isa{ancestor} is different to other definitions as it is an inductively defined relation rather than a recursive function, and we need further work to make this efficiently executable.
We characterise this relation with the following:

@@

Note that \isa{efficient\_ancestor\_simp} is tagged with a $\isa{code}$ ``attribute'', a hint that this theorem should be used as a transformation during code generation.
Isabelle/HOL takes this theorem as the definition, for code generation, of the \isa{efficient\_ancestor} relation and is able to directly translate this to a recursive function definition in each target backend---this transformation is ``semantics preserving'', as the characterisation was proved as a theorem within HOL.

Next, we demonstrate the correctness of our efficient definitions against our abstract definitions, and the properties that we prove are best described using an example, as they all follow a similar pattern.
We use the lemma \isa{efficient\_do\_op\_simulates} as our exemplar:

@@

We show that the \isa{efficient\_do\_op} and the \isa{do\_op} functions are in ``lock-step'', mapping related trees $t \preceq T$ to related trees $u \preceq U$ and producing identical logs of operations whenever they are applied to the same input operation, \isa{oper}.
This lock-step is repeated bottom-up for every efficient-abstract algorithm pairs, until we reach the final correctness theorem for \isa{apply\_ops}:

@@

Intuitively, this states that if both algorithms are applied to the same set of operations then they produce identical logs and also produce trees that contain the same set of parent-metadata-child records---i.e. the two trees are extensionally equivalent, despite having very different concrete representations.
Moreover, we have the following corollaries of the theorems \isa{apply\_ops\_commutes} and \isa{apply\_ops\_acyclic}, respectively, further establishing the correctness of our efficient algorithms:\footnote{The ``efficient'' analogue of \isa{apply\_ops\_unique\_parent} is trivially true, as the output tree is stored as a hash-map, and therefore too trivial to state as a theorem.}:

@@

Note the statement of \isa{efficient\_apply\_ops\_commutes} differs from that of \isa{apply\_ops\_commutes}---the latter asserts identical sets are obtained from convergence, whilst the former asserts that reading any key binding in $t$ gives the same result as a reading in $u$.
Whilst this statement appears weaker than that of \isa{apply\_ops\_acyclic}, it is correct, and the strongest statement one can hope for: the resulting hash-maps may contain the same key-value bindings but may not be structurally equal, depending on the ordering of operations.

@@

TODO: Coda~\cite{Kumar:1995wf}, Ficus~\cite{Reiher:1994wh}, Rumor~\cite{Guy:1999gy}, Roam~\cite{Ratner:1999fh}

>>>>

Isabelle/HOL can generate executable Haskell, OCaml, Scala, and Standard ML code from HOL type and constant definitions using a sophisticated code generation mechanism~\cite{DBLP:conf/flops/HaftmannN10}.
However,  care must be taken when using this mechanism, as:
\begin{enumerate}
\item
The mechanism is \emph{partial}: not all HOL definitions can be realised in executable form.
Definitions making essential use of choice principles (e.g.~our \isa{get\_parent}) are obvious examples of definitions that will cause problems.
However, \emph{inductively defined} relations characterised by a series of introduction rules, such as our \isa{ancestor} relation, are also problematic for the code generator as making them executable necessarily requires Prolog-like proof search and mode inference.
\item
Na\"{i}ve use of the code generator often leads to inefficient target language code.
This is especially true when using HOL's \isa{set} type, a type of potentially infinite mathematical sets.
Whilst convenient for theorem proving purposes, this type often leads to gross inefficiency in the extracted code.
\end{enumerate}

We must therefore first produce a series of refined definitions that are optimised for efficient execution rather than ease of theorem proving, before trying to evaluate our algorithms.
Within these definitions, we eschew the HOL \isa{set} type, and instead use a more efficient hash-map implementation provided by the Isabelle Collections Framework~\cite{DBLP:conf/itp/LammichL10}, a library of verified, efficient container datatypes and associated algorithms.
Write $('a::\{hashable\}, 'b)\ \isa{hm}$ for the type of hash-maps with key type $\isa{'a}$ and binding type $\isa{'b}$.
Note that keys have an explicit $\isa{hashable}$ constraint, meaning that this hash-map type may only be used with kay types with an associated hashing function.
With this type, we represent trees by the type $(\isa{'n}::\{hashable\}, \isa{'m} \times \isa{'n})\ \isa{hm}$, mapping from a \emph{child} node to its \emph{parent} node and associated metadata, if any such mapping exists.

We introduce an explicit \emph{simulation relation} between our two representations of trees.
Specifically, we say that a hash-map $\isa{t}$ of type $(\isa{'n}, \isa{'m} \times \isa{'n})\ \isa{hm}$ \emph{simulates} a set $T$ of type $(\isa{'n} \times \isa{'m} \times \isa{'n})\ \isa{set}$ whenever
\begin{displaymath}
\isa{hm.lookup}\ c\ t = \isa{Some}\ (m, p) \quad\text{iff}\quad (p, m, c) \in T \quad\text{forall $c$, $m$, and $p$}
\end{displaymath}
where $hm.lookup$ searches for a binding for key $\isa{c}$ in hash-map $t$, returning $\isa{Some}\ (m, p)$ when a binding from $c$ to $(m, p)$ exists in the hash-map, or $\isa{None}$ otherwise.
We write $t \preceq T$ to assert that the hash-map $t$ and set $T$ are contained within the simulation relation, defined above.
Note some useful corollaries follow immediately from this definition: e,g,~we have $t \preceq T$ implies $\isa{unique\_parent}\ \isa{T}$.

Our general strategy henceforth is to produce a new set of definitions---mirroring the structure of our existing definitions---wherein we replace all occurrences of the $(\isa{'n} \times \isa{'m} \times \isa{'n})\ \isa{set}$ type, and any associated set operations, with the $(\isa{'n}, \isa{'m} \times \isa{'n})\ \isa{hm}$ type and associated hash-map operations.
For the sake of clarity, we adopt the convention that any definition using hash-maps will be referred to as an ``efficient'' definition, and any definition making use of HOL's set type will be referred to as an ``abstract'' definition.
For most definitions this is a mechanical process, and we elide further description here.
However \isa{ancestor} is slightly different, as it is an inductively defined relation rather than a recursive function, and we therefore need further work to make this efficiently executable.
We characterise this relation recursively with the following theorem:

@@

Note that the theorem \isa{efficient\_ancestor\_simp} is tagged with a $\isa{code}$ ``attribute'', which is a hint that this theorem should be used as a transformation during the code generation process.
Isabelle/HOL takes this theorem as the definition, for code generation, of the \isa{efficient\_ancestor} relation and is able to directly translate this to a recursive function definition in each target backend.

Next, we demonstrate the correctness of our new efficient definitions against our older, abstract definitions.
The correctness properties that we prove are best described using an example, as they all follow a similar pattern.
We therefore focus on the lemma \isa{efficient\_do\_op\_simulates} which states:

@@

Note that we show that the \isa{efficient\_do\_op} and the \isa{do\_op} functions maintain a ``lock-step'' property, mapping related trees $t \preceq T$ to related trees $u \preceq U$ and producing identical logs of operations whenever they are applied to the same input operation, \isa{oper}.
This pattern of lock-step simulation is repeated bottom-up for every pair of efficient-abstract algorithm pairs, until we reach the final correctness theorem for \isa{apply\_ops}:

@@

This property captures our intuitive understanding of what it means for \isa{efficient\_apply\_ops} to be correct with respect to \isa{apply\_ops}: if both algorithms are applied to the same set of operations then they produce identical logs and also produce trees that contain the same set of parent-metadata-child records---i.e. the two trees are extensionally equivalent, despite having very different concrete representations.
We can further check that this theorem correctly captures the correctness of our \isa{apply\_ops} definition by recasting the various tree invariant and convergence theorems in terms of \isa{efficient\_apply\_ops}.
For example, we have the following corollaries of the theorems \isa{apply\_ops\_commutes} and \isa{apply\_ops\_acyclic}, respectively\footnote{The ``efficient'' analogue of \isa{apply\_ops\_unique\_parent} is trivially true, as the output tree is stored as a hash-map, and therefore too trivial to state as a theorem.}:

@@

Note the statement of \isa{efficient\_apply\_ops\_commutes} differs in an essential way from that of \isa{apply\_ops\_commutes}---the latter asserts that identical sets are obtained from convergence, whilst the former asserts that simply reading any key binding in $t$ gives the same result as reading in $u$.
Whilst this statement appears weaker than the statement of \isa{apply\_ops\_acyclic}, it is correct, as the resulting hash-maps may not have an identical representation in-memory---depending on the particular ordering of key-value insertions---despite containing the same key-value bindings.

@@

Unison is a file synchronisation tool with a formal specification~\cite{PierceVouillon:UnisonSpecTR} which allows two replicas, such as laptops, desktops or servers, to synchronise the state of a directory. 
Unison permits offline updates to both replicas. 
The (possibly divergent) state of two replicas are resynchronised by running the Unison tool on the two replicas. 
Unison works by recording a summary record of the state of the filesystem on each replica,  inspecting this summary whenever the unison program is invoked, and sharing any update information with the other replica in order to determine the changes made to the filesystems on each replica since they were last synchronised. 
This approach has the benefit that no special kernel extension or daemon process is required; instead Unison is an application program which inspects the current state of the filesystem whenever it is run. 
When presented with the move operations described in Figure~\ref{fig:move-same-item}, Unison duplicates the files, resulting in the outcome shown in Figure~\ref{fig:move-same-item}~(a). 
Unison is unable to automatically synchronise the move operations shown in Figure~\ref{fig:move-cycle} and instead asks the user to confirm whether changes made to directories A and B should be propagated or deleted. 
As a result, depending on user choice, both replicas can end up in one of four states, including those shown in Figure~\ref{fig:move-cycle}~(b), (c) or (d) or with an empty root directory on both replicas. 
In all cases, Unison successfully ensures both replicas are in the same state after synchronisation.

Coda is a distributed file system which allows clients to cache or \emph{hoard} copies of files otherwise stored in a shared data repository~\cite{kistler1992coda}. 
Clients can edit copies of data held in the cache while offline, during which time a kernel module keeps track of all updates. 
When the client comes back online, the client attempts to resynchronise changes with the repository. 
This design means that conflicts may arise if multiple clients edit the same files or directories while one or more clients are offline. 
One proposed solution to this is to support application-specific resolvers which "encapsulate the knowledge needed for file resolution"~\cite{Kumar:1995wf}, an approach is similar to that found in Bayou~\cite{Terry:1995dn}.

Ficus~\cite{Reiher:1994wh}, Rumor~\cite{Guy:1999gy}, Roam~\cite{Ratner:1999fh}
